---
title: "Functional analysis of single-cell transcriptomics"
author: "Pau Badia i Mompel"
date: "5/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Single-cell transcriptomics yield many molecular readouts that one by one are
hard to interpret. Functional analysis or "enrichment" tries to summarize this
information into interpretable concepts using prior knowledge, making it easier
to characterize the biological context being studied. 

The goals of this tutorial are to:

- Introduce the functional analysis concept
- Understand the differences in enrichment modeling
- Learn how to run functional analysis
- Interpret the obtained scores

To do so, we will analyse a downsampled data-set consisting of Blood myeloid
cells from healthy and COVID-19 infected patients available in the [Single Cell
Expression Atlas](https://www.ebi.ac.uk/gxa/sc/experiments/E-MTAB-9221/results/tsne?plotType=umap&plotOption=20).

> Silvin A, Chapuis N, Dunsmore G, Goubet AG, Dubuisson A et al. (2020) 
> Elevated Calprotectin and Abnormal Myeloid Cell Subsets Discriminate Severe 
> from Mild COVID-19.




```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```
